---
title: "Class 5: Data Viz with ggplot"
author: "Melanda (PID: A17473102)"
format: pdf
toc: true
---

## Background

There are lot's of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()` and tones of add-on packages like **ggplot2.**

For example here we make the same plot with both: 

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command 
`install.packages()`.

> **N.B** we never run an install cmd in a quearto code chunck or we will end up re-installing pacakges many many times - which is not what we want!

Every time we want to use one of these "add-on" packages we need to load it up in R with the `library()` function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things: 

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot
- The **geom**etry, the type of plot 

```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Stopping distance of old cars",
       subtitle = "Data from the `cars` object",
       x="Speed (mph)", 
       y="Distance (ft)")
```

```{r}
p + theme_bw()
```














## Gene expression plot 

We can read the input from the class website 

```{r}
url <- url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot 

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point()
```

Version 2 let's color by `State` so we can see the up and down significant genes compared ti all the "unchanging genes

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Version 3 plot, lets modify the default colors to something we like 

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  scale_colour_manual(values=c("purple", "pink", "blue")) + 
  labs(x="Control (no drug))", 
  y="Drug",
  title="Gene Expression Changes upon GLP-1 Drug") +
  theme_bw()
```


## Going Further

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

lets "facet" (i.e. make a separate plot) by continent rather than the big hot mess above. 

```{r}
ggplot(gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point(alpha=0.3) +
  facet_wrap(~continent)
```

## Custom plots 

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```

I want to "filter" down to a subsheet od this data. I will use the **dplyr** package to help me.

First I need to install it and then load it up.. 
`install.packages("dplyr")` and then `library(dplyr)`

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder, year==2007, country=="Ireland" )
```

```{r}
filter(gapminder, year==1977, country=="United States" )
```

```{r}
filter(gapminder, year==1977, country=="Ireland" )
```

```{r}
filter(gapminder, year==2007, country=="United States" )
```
> Q. Make a plot comparing 1977 and 2007 for all countries 

```{r}
input <- filter(gapminder, year %in% c(1977, 2007))
head(input)
```

```{r}
gapminder_1977 <- gapminder %>% filter(year==1977 | year==2007)

ggplot(gapminder_1977) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

